Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from ERA5-Land and MODIS

### 2024-01-12 article/paper_drought_Chile2023.qmd
avances articulo reunion 12 enero

@@

In this research, we present the dataset DDS4Chl, which provides climate variables and drought indices of water demand and  supply and vegetation productivity for continental Chile since 1981 at a monthly frequency. Those were gathered from the earth observation products ERA5-Land and MODIS. Then, we used DDS4Chl to analyze the impact of drought on different types of land cover classes in continental Chile. The specific objectives of the study are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of vegetation productivity (zcNDVI) with drought indices of water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

![(a) Chile and the five zones "norte grande", "norte chico",  "zona central", "zona sur", and "zona austral". (b) Topography reference map. (c) Land cover classes for 2021. (d) Persistent land cover classes (> 80%) for 2001-2021.](../output/figs/map_study_con_landcover.png){fig-pos='!ht'}

@@

For water supply and demand variabes, we used ERA5-Land [@Munoz2019], a reanalys dataset wich provides the evolution of land variables since 1950. It has a spatial resolution of 0.1°, hourly frequency and global coverage. We selected the variables for total precipitation, 2 meters temperature, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).

@@

##	Drought indices for water demand and supply, soil moisture and vegetation productivity

##	Analysis of a biomass proxy with drought indices of supply and demand

##	LULC change for 2001-2021 and its relation with water supply and demand

```{r LCprop}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone"
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
```

\begin{table}[!ht]
\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

@@

@@

##	LULC change for 2001-2021 and its relation with water supply and demand

>>>>

keep-tex: false

@@

In this research, we present the raster dataset DDS4Chl, which provides climate variables and drought indices of water demand and supply and vegetation productivity for continental Chile since 1981 at a monthly frequency. Those were gathered from the earth observation products ERA5-Land and MODIS. Then, we used DDS4Chl to analyze the impact of drought on different types of land cover classes in continental Chile. The specific objectives of the study are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of vegetation productivity (zcNDVI) with drought indices of water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop). The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009]. To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019]. "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb). In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna), followed by grassland (~16%), forest (8%), and croplands (5%). The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb). Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland.

```{r fig-studyArea}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral". (b) Topography reference map. (c) Land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022'
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/map_study_con_landcover.png")
```

@@

For water supply and demand variabes, we used ERA5-Land [@MunozSabater2021], a reanalys dataset wich provides the evolution of land variables since 1950. It has a spatial resolution of 0.1°, hourly frequency and global coverage. We selected the variables for total precipitation, 2 meters temperature, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).

@@

##	Impact fof water supply and demand, and soil moisture in vegetation productivity within landcover types

##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture

```{r fig-LCprop}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone"
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
```

@@

##	Impact for water supply and demand, and soil moisture in vegetation productivity

@@

##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture

\begin{table}[!ht]
\label{tab-landcoverTrend}
\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

\begin{table}[!ht]
\label{tab-corLandcover}
\caption{}
\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
\end{table}

### 2024-01-12 article/paper_drought_Chile2023.tex
avances articulo reunion 12 enero

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

In this research, we present the dataset DDS4Chl, which provides climate
variables and drought indices of water demand and~ supply and vegetation
productivity for continental Chile since 1981 at a monthly frequency.
Those were gathered from the earth observation products ERA5-Land and
MODIS. Then, we used DDS4Chl to analyze the impact of drought on
different types of land cover classes in continental Chile. The specific
objectives of the study are: i) to analyze the trend on multi-scalar
drought indices for water demand and supply, soil moisture, and
vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of vegetation productivity (zcNDVI)
with drought indices of water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

@@

\caption{(a) Chile and the five zones ``norte grande'', ``norte chico'',
``zona central'', ``zona sur'', and ``zona austral''. (b) Topography
reference map. (c) Land cover classes for 2021. (d) Persistent land
cover classes (\textgreater{} 80\%) for 2001-2021.}

@@

For water supply and demand variabes, we used ERA5-Land
\citep{Munoz2019}, a reanalys dataset wich provides the evolution of
land variables since 1950. It has a spatial resolution of 0.1°, hourly
frequency and global coverage. We selected the variables for total
precipitation, 2 meters temperature, and volumetric soil water layers
between 0 and 100cm of depth (layer 1 to layer 3).

@@

\hypertarget{drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
\subsection{Drought indices for water demand and supply, soil moisture
and vegetation
productivity}\label{drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}

\hypertarget{analysis-of-a-biomass-proxy-with-drought-indices-of-supply-and-demand}{%
\subsection{Analysis of a biomass proxy with drought indices of supply
and
demand}\label{analysis-of-a-biomass-proxy-with-drought-indices-of-supply-and-demand}}

\hypertarget{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand}{%
\subsection{LULC change for 2001-2021 and its relation with water supply
and
demand}\label{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand}}

@@

\caption{Proportion of land cover class from the persistent land cover
for 2001-2021 (\textgreater80\%) per macrozone}

@@

\begin{table}[!ht]
\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

@@

@@

@@

\hypertarget{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand-1}{%
\subsection{LULC change for 2001-2021 and its relation with water supply
and
demand}\label{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand-1}}

>>>>

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

In this research, we present the raster dataset DDS4Chl, which provides
climate variables and drought indices of water demand and supply and
vegetation productivity for continental Chile since 1981 at a monthly
frequency. Those were gathered from the earth observation products
ERA5-Land and MODIS. Then, we used DDS4Chl to analyze the impact of
drought on different types of land cover classes in continental Chile.
The specific objectives of the study are: i) to analyze the trend on
multi-scalar drought indices for water demand and supply, soil moisture,
and vegetation productivity for 1981--2023/2001--2023; ii) to evaluate
LULC change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of vegetation productivity (zcNDVI)
with drought indices of water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citep{Beck2023} and the landcover type
persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019}. ``Norte Grande'' and ``Norte
Chico'' predominate in an arid desert climate with hot (Bwh) and cold
(Bwk) temperatures. At the south of ``Norte Chico,'' the climate changes
to an arid steppe with cold temperatures (Bsk). Mainly, the land is
barren, with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
chilean matorral (shrubland and savanna), followed by grassland
(\textasciitilde16\%), forest (8\%), and croplands (5\%). The south part
of ``Centro'' and the north part of ``Austral'' are dominated by an
Oceanic climate (Cfb). Those zones are high in forest and grassland. The
southern part of the country has a tundra climate, and in Patagonia, it
is a cold semi-arid area with an extended surface of grassland.

@@

\caption{\label{fig-studyArea}(a) Chile with the Koppen-Geiger climate
classes and the five macrozones ``Norte Grande'', ``Norte Chico'',
``Centro'', ``Sur'', and ``Austral''. (b) Topography reference map. (c)
Land cover classes for 2022. (d) Persistent land cover classes
(\textgreater{} 80\%) for 2001-2022}

@@

For water supply and demand variabes, we used ERA5-Land
\citep{MunozSabater2021}, a reanalys dataset wich provides the evolution
of land variables since 1950. It has a spatial resolution of 0.1°,
hourly frequency and global coverage. We selected the variables for
total precipitation, 2 meters temperature, and volumetric soil water
layers between 0 and 100cm of depth (layer 1 to layer 3).

@@

\hypertarget{impact-fof-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}{%
\subsection{Impact fof water supply and demand, and soil moisture in
vegetation productivity within landcover
types}\label{impact-fof-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}}

\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
\subsection{LULC change for 2001-2022 and its relation with water supply
and demand, and soil
moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}

@@

\caption{\label{fig-LCprop}Proportion of land cover class from the
persistent land cover for 2001-2021 (\textgreater80\%) per macrozone}

@@

@@

\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity-1}{%
\subsection{Trend of drought indices for water demand/supply, soil
moisture, and vegetation
productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity-1}}

@@

\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
\subsection{Impact for water supply and demand, and soil moisture in
vegetation
productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}

@@

\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture-1}{%
\subsection{LULC change for 2001-2022 and its relation with water supply
and demand, and soil
moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture-1}}

\begin{table}[!ht]
\label{tab-landcoverTrend}
\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

\begin{table}[!ht]
\label{tab-corLandcover}
\caption{}
\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
\end{table}
